---
title: "Spotify Data Analysis"
author: 
  - Gadiel David Flores
thanks: "Code and data are available at: https://github.com/DavidFJ207/spotify_analysis"
date: today
date-format: long
abstract: "Using the `spotifyr` package, this paper analyzes audio data"
format: pdf
number-sections: true
bibliography: references.bib
---

# Introduction
This paper provides an analysis of music using the Spotify API. In this analysis, we examine three singers: Coldplay, Radiohead and The National. By utilizing Spotify’s extensive database of music and the Spotify API, we analysis three key audio features: Energy, danceability and loudness, which provide quantitative insights into the musical characteristics of each band. Overall, 

# Data {#sec-data}
For each artist, audio features of their tracks were downloaded, including `danceability`, `energy`, `valence`, and track duration.

```{r}
#| label: fig-data-summary
#| fig-cap: Data summary of audio features for Coldplay
#| echo: false
#| warning: false
#| message: false

# Load necessary libraries
library(ggplot2)
library(here)
library(dplyr)

# Load the data 
music <- read.csv(here::here("data/analysis_data/music_data.csv"))
# bar plot
ggplot(music, aes(x = Band, y = Energy, fill = as.factor(Energy))) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Energy",
       x = "Band",
       y = "Energy",
       fill = "Band") + 
  theme_minimal() +  
  scale_fill_grey(start = 0.8, end = 0.2) +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    plot.title = element_text(hjust = 0.5),  
    legend.position = "top"  
  )
```

## Data Overview

# Results {#sec-results}


```{r}
#| label: fig-danceability
#| fig-cap: Average Danceability for Coldplay, Radiohead, and The National
#| echo: false
#| warning: false
#| message: false

```

# Discussion {#sec-discussion}

## Trends {#sec-trends}

## Implications {#sec-implications}

# References {#sec-references}

For data analysis, we used R [@R-base] and the `spotifyr` package [@spotifyr]. The data was obtained through the Spotify API [@spotify-api], and the analysis follows standard practices in music data analysis [@music-data-analysis].