---
title: "Spotify Data Analysis"
subtitle: "Comparison of Audio Features Between Multiple Artists"
author: 
  - Gadiel David Flores, Tina Kim, Dannie Dai Dai, Yanfei Huang, Manjun Zhu
thanks: "Code and data are available at: https://github.com/DavidFJ207/spotify_analysis"
date: today
date-format: long
abstract: "Using the `spotifyr` package, this paper analyzes audio data features `danceability`, `energy`, and `loudness` by comparing between three artists, `Coldplay`, `Radiohead`, and `The National`."
format: pdf
number-sections: true
bibliography: references.bib
---
```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)

# Call the cleaned data
music_data <- 
  read_csv(file = here::here("data/analysis_data/music_data.csv"), 
           show_col_types = FALSE)

```

# Introduction
<<<<<<< HEAD
This paper provides an analysis of music using the Spotify API [@spotify-api]. In this analysis, we examine three singers: Coldplay, Radiohead and The National. By utilizing Spotify’s extensive database of music and the Spotify API, we analysis three key audio features: Energy, danceability and loudness, which provide quantitative insights into the musical characteristics of each band. Coldplay, Radiohead, and The National differ slightly in musical style: The National shows higher danceability, while Coldplay has greater energy and loudness than Radiohead.  

# Data {#sec-data}
For each artist, audio features of their tracks were downloaded from Spotify API [@spotify-api], including `danceability`, `energy`, and `loudness`.

```{r}
#| label: tbl-1
#| tbl-cap: Table of the cleaned dataset
#| echo: false

head(music_data, 3) |>
  kable(
    col.names = c("Band", "Danceability", "Energy", "Loudness"),
  booktabs = TRUE
  )
```

```{r}
#| label: fig-data-summary
#| fig-cap: Data summary of audio features for Coldplay
#| echo: false
#| warning: false
#| message: false

# Load the data 
music <- read.csv(here::here("data/analysis_data/music_data.csv"))
# bar plot Energy
ggplot(music, aes(x = Band, y = Energy, fill = as.factor(Energy))) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Energy",
       x = "Band",
       y = "Energy",
       fill = "Band") + 
  theme_minimal() +  
  scale_fill_grey(start = 0.8, end = 0.2) +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    plot.title = element_text(hjust = 0.5),  
    legend.position = "top"  
  )

# bar plot Loudness
ggplot(music, aes(x = Band, y = Loudness, fill = as.factor(Loudness))) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Loudness",
       x = "Band",
       y = "Loudness",
       fill = "Band") + 
  theme_minimal() +  
  scale_fill_grey(start = 0.8, end = 0.2) +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    plot.title = element_text(hjust = 0.5),  
    legend.position = "top"  
  )
```

## Data Overview

# Results {#sec-results}


```{r}
#| label: fig-danceability
#| fig-cap: Average Danceability for Coldplay, Radiohead, and The National
#| echo: false
#| warning: false
#| message: false

# Load the data 
music <- read.csv(here::here("data/analysis_data/music_data.csv"))

# bar plot Danceability
ggplot(music, aes(x = Band, y = Danceability, fill = as.factor(Danceability))) +
  geom_bar(stat = "identity", position = "dodge") +  
  labs(title = "Danceability",
       x = "Band",
       y = "Danceability",
       fill = "Band") + 
  theme_minimal() +  
  scale_fill_grey(start = 0.8, end = 0.2) +  
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),  
    plot.title = element_text(hjust = 0.5),  
    legend.position = "top"  
  )
```

## Trends {#sec-trends}
As bands get louder and the e
## Implications {#sec-implications}
The data highlights subtle but meaningful differences in the musical styles of Coldplay, Radiohead, and The National. **Coldplay**’s higher energy and loudness suggest a more mainstream, dynamic appeal, consistent with their stadium-filling sound and pop-rock influence. **Radiohead**'s lower energy and loudness reflect their more introspective, experimental approach, often favoring atmospheric, complex compositions over immediacy. **The National**'s higher danceability, combined with moderate energy and lower loudness, suggests a smoother, emotionally resonant style, balancing accessibility with depth. These variations imply how each band resonates with their respective audiences, shaping listener experiences through distinct blends of energy, mood, and engagement.
# References {#sec-references}

For data analysis, we used R [@R-base] and the `spotifyr` package [@spotifyr]. The data was obtained through the Spotify API [@spotify-api], and the analysis follows standard practices in music data analysis [@music-data-analysis].
